Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-12 TOMS_Paper/assembly.tex
Some other tweaks + captions

\begin{figure}[t]
\tiny
\centering     
\subfigure[Factorized, showing zeros, code]{\label{fig:withzeros-code}\lstinputlisting{listings/withzeros.code}}
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subfigure[Factorized, skipping zeros, code]{\label{fig:withzeros-skipped-code}\lstinputlisting{listings/skipzeros.code}}
\caption{Without and with skipping zeros}\label{fig:skip-code}
\end{figure}

Consider Figure~\ref{fig:withzeros-code}, which is an enriched version of the Burgers excerpt in Figure~\ref{fig:invariant-code}. The code is instantiated for the specific case of polynomial order 1, Lagrange elements on a 2D mesh. The array \texttt{D} represents a tabulated derivative of a basis function at the various quadrature points. There are four zero-valued columns. Any multiplications or additions along these columns could (should) be skipped to avoid irrelevant floating point operations. The solution adopted in~\cite{quadrature1} is not to generate the zero-valued columns (i.e. to generate a dense 6$\times$2 array), to reduce the size of the iteration space over test and trial functions (from 6 to 2), and to use an indirection array (e.g. $ind = \lbrace 3, 5\rbrace$) to update the right entries in the element tensor $A$. This prevents, among the various optimizations, effective SIMD vectorization, because memory loads and store would eventually reference non-contiguous locations. 
%on the other hand, reshrinki tutto su un array solo...

Our approach is based on using domain knowledge and symbolic execution. We discern the origin of zero-valued columns: for example, those due to taking derivatives on the reference element from those inherent to using mixed (vector) elements. In the running Burgers example, the use of vector function spaces require the generation of a zero-block (columns 0, 1, 2 in the array \texttt{D}) to correctly evaluate the local element matrix while iterating along the space of test and trial functions. The two key observations are that 1) the number of zero-valued columns caused by using vector function spaces is, often, much larger then that due to derivatives, and 2) such columns are contiguous in memory. Based on this, we aim at avoiding iteration only along the block of zero-valued columns induced by mixed (vector) elements.

@@

We achieve our goal by symbolic execution. The Expression Rewriter expects some indication in the input exposing the location of the zero-valued columns due to mixed (vector) function spaces, in each tabulated basis function. This indication comes either in the form of code annotation, if the input to COFFEE were provided as pure C (as shown in Figure~\ref{fig:withzeros-code}), or by suitably decorating basis functions nodes, if the input were an abstract syntax tree. Then, the rewrite rules are applied and each statement is executed symbolically. For example, consider the assignment \texttt{T2[r] = d*D[i][k]+e*E[i][k]} in Figure~\ref{fig:withzeros-code}. Array \texttt{D} has non-zero-valued columns in positions $NZ_D = [3,5]$; we also assume array \texttt{E} has non-zero-valued columns in positions $NZ_E = [0,2]$. Multiplications by scalar values do not affect the propagation of non-zero-valued columns. On the other hand, when symbolically executing the sum of the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we track that the target identifier \texttt{T2} will have non-zero-valued columns in positions $NZ_D \cup NZ_E = [0,5]$. Eventually, exploiting the $NZ$ information computed and associated with each identifier, we split the original assembly expression into multiple sets of sub-expressions, each set characterized by the same range of non-zero-valued columns. In our example, assuming that $NZ_{T1} = [3,5]$ and $NZ_A = [3,5]$, there are two of such sets, this leads to the generation of two distinct iteration spaces, one for each set, as in Figure~\ref{fig:withzeros-skipped-code}.

\section{Code Specialization}
\label{sec:code-spec}
Code specialization provides a range of code transformations tailored to optimizing for instruction-level parallelism and register locality. As summarized in Section~\ref{sec:summary-opts} and described in~\cite{Luporini}, padding and data alignment, expression splitting, and outer-product vectorization, which is a domain-aware implementation of vector-register tiling, are examples of optimizations that the Code Specializer is capable of applying. In this paper, we enrich this set of transformations as explained in the following sections.

\subsection{Padding and Data Alignment Revisited}
\label{sec:padding}
Padding and data alignment as described in~\cite{Luporini} must be refined for the case in which computation over zero-valued columns is avoided. We recall effective SIMD vectorization can be achieved only if the innermost loop size is a multiple of the vector register length. In the case of an AVX instruction set, for example, we want the size of innermost loops to be multiples of 4 (vector registers can fit up to four double-precision floats). Moreover, efficient loads and stores can be issued only if all of the involved arrays' base addresses are multiple of the vector register length.

Consider again the code in Figure~\ref{fig:withzeros-skipped-code}. The arrays in the loop nest $[j1,k1]$ can be padded and the right bound of loop \texttt{k1} can be safely increased to 8: eventually, values computed in the region $M[0:3][6:8]$ will be discarded. Then, by explicitly aligning arrays and using suitable pragmas (e.g. \texttt{$\#$pragma simd} for the Intel compiler), effective SIMD auto-vectorization can be obtained.

There are some complications in the cose of loops $[j0,k0]$. Here, increasing the loop bound to 4, despite being still safe, has no effect: for instance, the starting addresses of load instructions are \texttt{T1[3]} and \texttt{A[i][3]}, which are not aligned. One solution is to start iterating from the closest index that would ensure data alignment: in this specific case, $k0=0$, which would mean losing the effect of expression rewriting. Another possibility is to attain to non-aligned accesses. COFFEE can generate code for both situations, so we leave the autotuner, described in Section~\ref{sec:autotune}, in charge of determining the optimal transformation.

Note that in some cases it is not even safe to increase loop bounds, since extra iterations would access non-zero entries in the local element matrix $M$. In such a situation, there may be no possibility of restoring data alignment at all.

\subsection{Exposing Linear Alegbra Operations}
In~\cite{Luporini}, we introduced the idea of transforming the element matrix evaluation into a sequence of calls to highly-optimized dense matrix-matrix multiply routines (henceforth DGEMM), for instance MKL or ATLAS BLAS. We compared COFFEE's optimizations with hand-written MKL-based kernels, showing how the small sizes of the involved arrays impair DGEMM routines, which are usually tuned for large arrays. The study was conducted only in the case of a specific Helmholtz problem. There are scenarios, however, in which tabulated basis functions sizes grow up to a point for which turning to BLAS may actually lead to better performance. For example, this can happen when relatively-high polynomial orders are used or if coefficients are present in the form, since they are both responsible for the size of tabulated basis functions. To explore these scenarios, we have developed an algorithm that reduces any assembly expression evaluation to a sequence of DGEMM calls.

Here, we informally provide the main steps of the algorithm. By fully applying the rewrite rules in Figure~\ref{fig:rewrite-rules}, an assembly expression is reduced to a summation, over each quadrature point, of outer products along the test and trial functions. Each outer product is then isolated, i.e. the assembly expression is split into chunks, each chunk representing an outer product over test and trial functions. Statements in the bodies of the surrounding loops (e.g. coefficients evaluation at a quadrature point, temporaries introduced by expression rewriting) are vector-expanded and hoisted completely outside of the loop nest, similarly to what we have described in Section~\ref{sec:expr-rewriter-obj}. This renders perfect the loop nest; that is, there is no intervening code among the various loops. The element matrix evaluation has now become a sequence of dense matrix-matrix multiplies (transposition aside)
\begin{equation*}
A_{jk} = \sum_{i} x_{0_{ij}} \cdot y_{0_{ik}} + \sum_{i} x_{1_{ij}} \cdot y_{1_{ik}} + ...
\end{equation*}
where $x0, x1, y0, y1, ...$ are tabulated basis functions or vector-expanded temporaries introduced at expression rewriting time. Eventually, the storage layout of the involved operands is changed so as to be conforming to the BLAS interface (e.g. two dimensional arrays are flatten as one dimensional arrays). The translation into a sequence of DGEMM calls is the last, straightforward step.

@@

In~\cite{Luporini}, we proposed a simple cost model to select, for a given a problem, the optimal combination of transformations. In this paper, we have added a significant number of options to the set of possible optimizations, so the selection problem is now far more challenging. The sole Expression Rewriter, for instance, could generate many possible code variants by applying rewrite rules to different extents. We found difficult extending the cost model to effectively cover such a large search space.

We tackle the optimization selection problem by compiler autotuning. Not only does it allow to determine the best combination of transformations out of the set presented so far, also it enables exploring parametric low-level optimizations, such as loop unroll, unroll-and-jam, and interchange, by trying different unroll factors and loop permutations. By leveraging the cost model defined in our previous study, domain-awareness, and a set of heuristics, we manage to keep the autotuner overhead at a minimum, whilst achieving significant speed ups over the purely cost-model-based implementation. In particular, our autotuner usually requires order of seconds to determine the fastest kernel implementation, a negligible overhead when it comes to iterate over real-life unstructured meshes, which can contain up to trillions of elements (e.g. ~\cite{Rossinelli2013}).

COFFEE analyzes the input problem and decides what variants it is worth testing, as described later. Each variant is obtained by requesting specific transformations to the Expression Rewriter and the Code Specializer. The possible variants are then provided to the autotuner, in the form of abstract syntax trees. The autotuner is a template-based code generator. By inspecting an abstract syntax tree, it determines how to generate ``wrapping'' code that 1) initializes kernel's input variables with fictitious values and 2) calls the kernel. These two points are executed repeatedly in a \emph{while} loop for a pre-established amount of time (order of milliseconds). At the exit of the \emph{while} loop, the times the kernel was invoked is recorded. Eventually, the variant executed the largest number of iterations is designated as the fastest implementation. Suitable compiler directives are used to prevent inlining of all function calls: this avoids the situation in which some variants are inlined and some are not, which would fake the autotuner's output.

The autotuning process is dynamic: depending on the complexity of the input problem, more or less variants are tried. General heuristics, which can be considered a revisited version of those presented in~\cite{nek5000}, are applied
\begin{itemize}
\item Loop permutations that are likely to worsen the performance are excluded from the search space. According to the cost model, and for the same reasons explained in~\cite{Luporini}, we enable only variants in which the loop over quadrature points is either the outermost or the innermost. This is due to the fact that versions of the code in which such loop lies between the test and trial functions loops are typically lower performing.
\item The unroll factors must divide the loop bounds evenly to avoid the introduction of reminder (scalar) loops.
\item The innermost loop is never explicitly unrolled. This is because we expect auto-vectorization along this loop, so memory accesses should be kept unit-stride. 
\end{itemize}
The autotuner is also domain-aware: the following heuristics, which capture properties of the computational domain, are exploited
\begin{itemize}
\item The lengths of test and trial functions loops are identical in some cases, for example when they originate from the same function space. In such cases, since for the employed storage layout the memory access pattern is symmetrical along these two loops, we prune their interchange from the search space. 
\item The larger is the polynomial order of the method, the larger is the assembly loop nest. In these cases, we impose a bound on the loop nest's overall unroll factor (which we found empirically) to avoid uselessly testing too many unroll factors. 
\item On the other hand, if the polynomial order is low, i.e. when the loop nest is small, we prune variants that we know will be low-performing, e.g. those resorting to BLAS.
\item We select two levels of expression rewriting. In the ``base'' level, only generalized loop-invariant code motion, as described in~\cite{Luporini}, is applied. This means that only a subset of the rewrite rules exposed in~\ref{fig:rewrite-rules} will be considered. In the ``aggressive'' level, all of the rewrite rules are applied. Many other trade-offs, which we do not explore, would be feasibile, however. 
\item For the expression splitting optimization described in~\cite{Luporini} and summarized in Section~\ref{sec:summary-opts}, we test only three split factors, namely 1, 2, 4. Also, if the input problem uses mixed function spaces, the iteration space is already split by the Expression Rewriter to avoid computation over zero-columns; in these cases, we do not further apply expression splitting.
\item Based on the cost model, the padding and data alignment optimization is always applied. 
\end{itemize}

%COFFEE allows the user to abstract from finding the optimal optimization strategy for a given problem by resorting to its model-driven, dynamic autotuner.

@@

We analyze the run-time performance of four fundamental real problems, which comprise the differential operators that are most common in finite element methods. In particular, our study includes problems based upon the Helmholtz and Poisson equations, as well as elasticity- and hyperelasticity-like forms. The Unified Form Language~\cite{UFL} specification for these forms, which is the domain specific language that both Firedrake and FEniCS use to express weak variational form, is available at~\cite{Code-location}.

@@

The values that we report include the cost of local assembly as well as the cost of matrix insertion. However, the unstructured mesh has been made small enough to fit the L3 cache, so as to minimize the noise due to any operations that are not part of the element matrix evaluation itself. It has been discussed in many places already (e.g.~\cite{quadrature1}) that as the complexity of a form increseas, the cost of local assembly becomes dominant. All codes were executed in the context of the Firedrake framework.

We do not compare to the FEniCS Form Compiler's tensor contraction mode~\cite{Kirby-FEM-opt} because of three reasons: first, in~\cite{quadrature1} it has been demonstrated the superiority of quadrature as the complexity of a form increases, so it would be superfluous to repeat the same analysis. Second, our aim is to show the effect of low-level optimizations on the code, especially SIMD vectorization, which is not feasible in tensor mode. Third, tensor mode code generation fails due to hardware limitations in many of the test cases that we show below.

We vary several aspects of each form, which follows the approach and the notation of~\cite{quadrature1} and~\cite{Francis}
\begin{itemize}
\item The polynomial order of basis functions, $q \in \lbrace1, 2, 3, 4\rbrace$
\item The polynomial order of coefficient (also called pre-multiplying) functions, $p \in \lbrace1, 2, 3, 4\rbrace$
\item The number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
On the other hand, other aspects are fixed 
\begin{itemize}
\item The space of both basis and coefficient functions is Lagrange
\item The mesh is three-dimensional, made of tetrahedrons, for a total of 4374 cells
\end{itemize}

Each figure reported in the following corresponds to one specific problem and must be read as a grid of plots. Each grid has two axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis. The top-left plot will show values for $[q=1, p=1]$; the plot on its right will be $[q=1, p=2]$ and so on. The diagonal of the grid shows plots for which both basis and coefficient functions originate from the same function space. Therefore, a grid can be read in many different ways, which allows us to make structured considerations on the effect of the various optimizations.

@@

We first comment on results of general applicability. We note there is a trend of COFFEE's optimizations to become more and more effective as $q$, $p$, and $nf$ increase. This is because most of the transformations applied aim at optimizing for arithmetic intensity and SIMD vectorization, which obviously have a strong impact when arrays and iteration spaces are large. The corner cases of this phenomenon are indeed $[q=1, p=1]$ and $[q=4, p=4]$. We also observe how \texttt{auto}, in almost all scenarios, outperforms all of the other variants. In particular, it is not a surprise that \texttt{auto} is faster than \texttt{fix}, since \texttt{fix} is one of the autotuner's tested variants, as explained in Section~\ref{sec:autotune}. This proves the quality of the work presented in this paper, which shows significant advances over~\cite{Luporini}. The reasons for which \texttt{auto} exceeds both original code and \texttt{ffc} are discussed for each specific problem next. Also, details on the ``optimal'' code variant determined by autotuning are given in Section~\ref{sec:auto-analysis}.

\paragraph{Helmholtz}
\begin{figure}[t]  
\includegraphics[scale=0.7]{perf-results/helmholtz}
\caption{Helmholtz results.}\label{fig:helmholtz}
\end{figure}
The results for the Helmholtz problem are provided in Figure~\ref{fig:helmholtz}. We observe that \texttt{ffc} slows the code down, especially for $q \geq 3$, as a consequence of using indirection arrays in the generated code that, as explained in Section~\ref{sec:zeros}, prevent, among the other compiler's optimizations, SIMD auto-vectorization. The \texttt{auto} version results in minimal performance improvements over \texttt{fix} when $nf=0$, unless $q=4$. This is due to the fact that if the loop over quadrature points is relatively small, then close-to-peak performance is obtainable through basic expression rewriting and code specialization; in this circumstance, generalized loop-invariant code motion and padding plus data alignment. The trend changes dramatically as $nf$ and $q$ increase: a more ample spectrum of transformations must be considered to find the optimal local assembly implementation. We will comment on this in details in the next section.

\paragraph{Elasticity}
\begin{figure}[t]  
\includegraphics[scale=0.7]{perf-results/elasticity}
\caption{Elasticity results.}\label{fig:elasticity}
\end{figure}
Figure~\ref{fig:elasticity} illustrates results for the Elasticity problem. This form uses vector-valued spaces for the basis functions, so here transformations avoiding computation over zero-valued columns are of key importance. The \texttt{ffc} set of optimizations leads to notable improvements over the original code at $q=1$. The use of inderection arrays allows to phisically eliminate zero-valued columns at code generation time; as a consequence, different tabulated basis functions are merged into a single array. Therefore, despite the execution being purely scalar because of indirection arrays, the reduction in arithmetic intensity and register pressure imply improvement in performance. Nevertheless, \texttt{auto} remains in general the best choice, with gains over \texttt{ffc} that are wider as $p$ and $nf$ increase.

For $q \geq 2$, in \texttt{ffc} the lack of SIMD vectorization counterbalances the decrease in the number of floating point operations, leading to speed ups over the original code that only occasionally exceeds 1$\times$. On the other hand, the successful application of the zero-avoidance optimization while preserving code specialization plays a key role for \texttt{auto}, resulting in much higher performance code especially at $q=2$ and $q=3$.

@@

Speed ups for the hyperelasticity form are shown in Figure~\ref{fig:hyperelasticity}. Experiments for $nf \geq 3$ could not be executed because of FEniCS-Form-Compiler's technical limitations.

>>>>

\begin{figure}[t]
\tiny
\centering     
\subfigure[Original (simplified) code. Note the annotation over the definition of the tabulated basis function \texttt{D}, which is used to identify the presence of zero-valued columns]{\label{fig:withzeros-code}\lstinputlisting{listings/withzeros.code}}
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subfigure[The code after symbolic execution took place]{\label{fig:withzeros-skipped-code}\lstinputlisting{listings/skipzeros.code}}
\caption{Simplified excerpt of local assembly code from a Burgers form using vector-valued basis functions, before and after symbolic execution is performed to rewrite the iteration space}\label{fig:skip-code}
\end{figure}

Consider Figure~\ref{fig:withzeros-code}, an enriched version of the Burgers excerpt in Figure~\ref{fig:invariant-code}. The code is instantiated for the specific case of polynomial order 1, Lagrange elements on a 2D mesh. The array \texttt{D} represents a tabulated derivative of a basis function at the various quadrature points. There are four zero-valued columns. Any multiplications or additions along these columns could (should) be skipped to avoid irrelevant floating point operations. The solution adopted in~\cite{quadrature1} is not to generate the zero-valued columns (i.e. to generate a dense 6$\times$2 array), to reduce the size of the iteration space over test and trial functions (from 6 to 2), and to use an indirection array (e.g. $ind = \lbrace 3, 5\rbrace$) to update the right entries in the element tensor $A$. This prevents, among the various optimizations, effective SIMD vectorization, because memory loads and store would eventually reference non-contiguous locations. 
%on the other hand, reshrinki tutto su un array solo...

Our approach is based on using domain knowledge and symbolic execution. We discern the origin of zero-valued columns: for example, those due to taking derivatives on the reference element from those inherent to using mixed (vector) elements. In the running Burgers example, the use of vector function spaces require the generation of a zero-block (columns 0, 1, 2 in the array \texttt{D}) to correctly evaluate the local element matrix while iterating along the space of test and trial functions. The two key observations are that 1) the number of zero-valued columns caused by using vector function spaces is, often, much larger then that due to derivatives, and 2) such columns are contiguous in memory. Based on this observation, we aim at avoiding iteration only along the block of zero-valued columns induced by mixed (vector) elements.

@@

We achieve our goal by means of symbolic execution. The Expression Rewriter expects some indication about the location of the zero-valued columns induced by mixed (vector) function spaces, for each tabulated basis function. This indication comes either in the form of code annotation, if the input to COFFEE were provided as pure C (as shown in Figure~\ref{fig:withzeros-code}), or by suitably decorating basis functions nodes, if the input were an abstract syntax tree. Then, the rewrite rules are applied and each statement is executed symbolically. For example, consider the assignment \texttt{T2[r] = d*D[i][k]+e*E[i][k]} in Figure~\ref{fig:withzeros-code}. Array \texttt{D} has non-zero-valued columns in the range $NZ_D=[3,5]$; we also assume array \texttt{E} has non-zero-valued columns in the range $NZ_E=[0,2]$. Multiplications by scalar values do not affect the propagation of non-zero-valued columns. On the other hand, when symbolically executing the sum of the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we track that the target identifier \texttt{T2} will have non-zero-valued columns in the range $NZ_E \| NZ_D=[0,5]$. Eventually, exploiting the $NZ$ information computed and associated with each identifier, we split the original assembly expression into multiple sets of sub-expressions, each set characterized by the same range of non-zero-valued columns. In our example, assuming that $NZ_{T1}=[3,5]$ and $NZ_A=[3,5]$, there are two of such sets, which leads to the generation of two distinct iteration spaces (one for each set), as in Figure~\ref{fig:withzeros-skipped-code}.

\section{Code Specialization}
\label{sec:code-spec}
Code specialization provides a range of code transformations tailored to optimizing for instruction-level parallelism and register locality. As summarized in Section~\ref{sec:summary-opts} and described in~\cite{Luporini}, padding and data alignment, expression splitting, and outer-product vectorization, which is a domain-aware implementation of vector-register tiling, are examples of optimizations that the Code Specializer is capable of applying. In this paper, we revisit and enrich this set of transformations as explained in the following sections.

\subsection{Padding and Data Alignment Revisited}
\label{sec:padding}
Padding and data alignment as described in~\cite{Luporini} must be refined for the case in which computation over zero-valued columns is avoided. We recall effective SIMD (auto-)vectorization can be achieved only if the innermost loop size is a multiple of the vector register length, in which case the compiler needs not to introduce a remainder scalar loop. In the case of an AVX instruction set, for example, we want to round the size of the innermost loops to the closest multiple of 4 (AVX registers can fit up to four double-precision floats). This can be done provided that arrays are padded, if necessary. Moreover, loads and stores instructions are efficient only if their addresses are suitably aligned to cache boundaries; this is obtainable by enforcing the base address of each padded array to be a multiple of the vector length.

Consider again the code in Figure~\ref{fig:withzeros-skipped-code}. The arrays in the loop nest \texttt{[j1,k1]} can be padded and the right bound of loop \texttt{k1} can be safely increased to 8: eventually, values computed in the region \texttt{M[0:3][6:8]} will be discarded. Then, by explicitly aligning arrays and using suitable pragmas (e.g. \texttt{$\#$pragma simd} for the Intel compiler), effective SIMD auto-vectorization can be obtained for this loop nest.

There are some complications in the case of loops \texttt{[j0,k0]}. Here, increasing the loop bound to 4 is still safe, assuming that both \texttt{T1} and \texttt{A} are padded with zero-valued entries, but it has no effect: the starting addresses of the load instructions would be \texttt{T1[3]} and \texttt{A[i][3]}, which are not aligned. One solution is to start iterating from the closest index that would ensure data alignment: in this specific case, $k0=0$. However, this would imply losing (partially in general, totally for this loop nest) the effect of the zero-avoidance transformation. Another possibility is to attain to non-aligned accesses. COFFEE can generate code for both situations, so we leave the autotuner, described in Section~\ref{sec:autotune}, in charge of determining the optimal transformation.

Note that in some circumstances the previous solution cannot be applied, since extra iterations might end up accessing non-zero entries in the local element matrix \texttt{M}. In such a situation, there is no possibility of recovering data alignment.

\subsection{Exposing Linear Alegbra Operations}
In~\cite{Luporini}, we introduced the idea of transforming the element matrix evaluation into a sequence of calls to highly-optimized dense matrix-matrix multiply routines (henceforth DGEMM), for instance MKL or ATLAS BLAS. We compared COFFEE's optimizations with hand-written MKL-based kernels, showing how the small sizes of the involved arrays impair DGEMM routines, which are usually tuned for large arrays. The study was conducted only in the case of a specific Helmholtz form. There are scenarios, however, in which tabulated basis functions sizes grow up to a point for which turning to BLAS may actually lead to better performance. For example, this can happen when relatively-high polynomial orders are used or if coefficients are present in the form, since they are both responsible for the size of basis functions tabulated at quadrature points. To explore these scenarios, we have developed an algorithm that reduces any assembly expression evaluation to a sequence of DGEMM calls.

We informally provide the main steps of the algorithm. By fully applying the rewrite rules in Figure~\ref{fig:rewrite-rules}, an assembly expression is reduced to a summation, over each quadrature point, of outer products along the test and trial functions. Each outer product is then isolated, i.e. the assembly expression is split into chunks, each chunk representing an outer product over test and trial functions. Statements in the bodies of the surrounding loops (e.g. coefficients evaluation at a quadrature point, temporaries introduced by expression rewriting) are vector-expanded and hoisted completely outside of the loop nest, similarly to what we have described in Section~\ref{sec:expr-rewriter-obj}. This renders the loop nest perfect; that is, there is no intervening code among the various loops. The element matrix evaluation has now become a sequence of dense matrix-matrix multiplies (transposition aside)
\begin{equation*}
A_{jk} = \sum_{i} x_{0_{ij}} \cdot y_{0_{ik}} + \sum_{i} x_{1_{ij}} \cdot y_{1_{ik}} + ...
\end{equation*}
where $x0, x1, y0, y1, ...$ are tabulated basis functions or vector-expanded temporaries introduced at expression rewriting time. Eventually, the storage layout of the involved operands is changed so as to be conforming to the BLAS interface (e.g. two dimensional arrays are flatten as one dimensional arrays). The translation into a sequence of DGEMM calls is the last, straightforward step.

@@

In~\cite{Luporini}, we proposed a simple cost model to drive the optimization process. In this paper, we have added a significant number of options to the set of possible optimizations, so the selection problem is now far more challenging. The sole Expression Rewriter, for instance, can generate a wide variety of implementations by just applying rewrite rules to different extents. We found difficult extending the cost model to effectively cover such a large search space.

We tackle the optimizations selection problem by compiler autotuning. Not only does it allow to determine the best combination of transformations, also it enables exploring parametric low-level optimizations, such as loop unroll, unroll-and-jam, and interchange, by trying different unroll factors and loop permutations. By leveraging the cost model defined in our previous study, domain knowledge, and a set of heuristics, we manage to keep the autotuner overhead at a minimum, whilst achieving significant speed ups over the purely cost-model-based implementations in all of the forms we have evaluated. In particular, our autotuner usually requires order of seconds to determine the fastest kernel implementation, a negligible overhead when it comes to iterate over real-life unstructured meshes, which can contain up to trillions of elements (e.g. ~\cite{Rossinelli2013}).

COFFEE analyzes the input problem and decides what variants it is worth testing through autotuning, as described later. Each variant is obtained by requesting specific transformations to the Expression Rewriter and the Code Specializer. The possible variants are then provided to the autotuner, in the form of abstract syntax trees. The autotuner is a template-based code generator. By inspecting an abstract syntax tree, it determines how to generate ``wrapping'' code that 1) initializes kernel's input variables with fictitious values and 2) calls the kernel. These two points are executed repeatedly in a \emph{while} loop for a pre-established amount of time (order of milliseconds). At the exit of the \emph{while} loop, the times the kernel was invoked is recorded. Eventually, the variant executed the largest number of iterations is designated as the fastest implementation. Suitable compiler directives are used to prevent inlining of all function calls: this avoids the situation in which some variants are inlined and some are not, which would fake the autotuner's output.

The autotuning process is dynamic: depending on the complexity of the input problem, more or less variants are tried. General heuristics, which can be considered a revisited version of those presented in~\cite{nek5000}, are applied
\begin{itemize}
\item Loop permutations that are likely to worsen the performance are excluded from the search space. According to the cost model, and for the same reasons explained in~\cite{Luporini}, we enable only variants in which the loop over quadrature points is either the outermost or the innermost. This is due to the fact that versions of the code in which such loop lies between the test and trial functions loops are typically lower performing.
\item The unroll factors must divide the loop bounds evenly to avoid the introduction of reminder (scalar) loops.
\item The innermost loop is never explicitly unrolled. This is because we expect auto-vectorization along this loop, so memory accesses should be kept unit-stride. 
\end{itemize}
The autotuner is also domain-aware: the following heuristics, which capture properties of the computational domain, are exploited
\begin{itemize}
\item The lengths of test and trial functions loops are identical in some cases, for example when they originate from the same function space. In such cases, since for the employed storage layout the memory access pattern is symmetrical along these two loops, we prune their interchange from the search space. 
\item The larger is the polynomial order of the method, the larger is the assembly loop nest. In these cases, we impose a bound on the loop nest's overall unroll factor (which we found empirically) to avoid uselessly testing too many unroll factors. 
\item On the other hand, if the polynomial order is low, i.e. when the loop nest is small, we prune variants that we know will be low-performing, e.g. those resorting to BLAS.
\item We specifically select two levels of expression rewriting. In the ``base'' level, only generalized loop-invariant code motion, as described in~\cite{Luporini}, is applied. This means that only a subset of the rewrite rules exposed in~\ref{fig:rewrite-rules} will be considered. In the ``aggressive'' level, all of the rewrite rules are applied. Many other trade-offs, which we do not explore, would be feasibile, however. 
\item For the expression splitting optimization described in~\cite{Luporini} and summarized in Section~\ref{sec:summary-opts}, we test only three split factors, namely 1, 2, 4. Also, if the input problem uses mixed function spaces, the iteration space is already split at expression rewriting time to avoid computation over zero-valued columns; in these cases, we do not further apply expression splitting.
\item Based on the cost model, the padding and data alignment optimization is always applied. 
\end{itemize}

All of the previous points contribute to minimize the overhead of the autotuning process. We will discuss the actual cost of autotuning in Section~\ref{sec:auto-analysis}.
%COFFEE allows the user to abstract from finding the optimal optimization strategy for a given problem by resorting to its model-driven, dynamic autotuner.

@@

We analyze the run-time performance of four fundamental real problems, which comprise the differential operators that are most common in finite element methods. In particular, our study includes problems based upon the Helmholtz and Poisson equations, as well as elasticity- and hyperelasticity-like forms. The Unified Form Language~\cite{UFL} specification for these forms, which is the domain specific language that both Firedrake and FEniCS use to express weak variational form, is available at~\cite{ufl-code}.

@@

The values that we report include the cost of local assembly as well as the cost of matrix insertion. However, the unstructured mesh has been made small enough to fit the L3 cache, so as to minimize the ``noise'' due to any operations that are not part of the element matrix evaluation itself. However, it has been reiterated over and over (e.g.~\cite{quadrature1}) that as the complexity of a form increseas, the cost of local assembly becomes dominant. All codes were executed in the context of the Firedrake framework.

We do not compare to the FEniCS Form Compiler's tensor contraction mode~\cite{Kirby-FEM-opt} because of three reasons: first, in~\cite{quadrature1} it has been demonstrated the superiority of quadrature as the complexity of a form increases, so it would be superfluous to repeat the same analysis. Second, our aim is to show the effect of low-level optimizations on the code, especially SIMD vectorization, which is not feasible in tensor mode due to the inherent structure of the code. Third, tensor mode code generation fails due to hardware limitations in many of the test cases that we show below.

We vary several aspects of each form, which follows the approach and the notation of~\cite{quadrature1} and~\cite{Francis}
\begin{itemize}
\item The polynomial order of basis functions, $q \in \lbrace1, 2, 3, 4\rbrace$
\item The polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
\item The number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
On the other hand, other aspects are fixed 
\begin{itemize}
\item The space of both basis and coefficient functions is Lagrange
\item The mesh is three-dimensional, made of tetrahedrons, for a total of 4374 cells
\end{itemize}

Figures~\ref{fig:helmholtz},~\ref{fig:elasticity},~\ref{fig:poisson}, and~\ref{fig:hyperelasticity}, which will be deeply commented in the next section, must be read as ``plots, or grids, of plots''. Each grid (figure) has two logical axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis. The top-left plot in a grid shows speed ups for $[q=1, p=1]$; the plot on its right does the same for $[q=1, p=2]$, and so on. The diagonal of the grid shows plots for which basis and coefficient functions have same polynomial order, that is $q=p$. Therefore, a grid can be read in many different ways, which allows us to make structured considerations on the effect of the various optimizations.

@@

We first comment on results of general applicability. By looking at the various figures, we note there is a trend in COFFEE's optimizations to become more and more effective as $q$, $p$, and $nf$ increase. This is because most of the transformations applied aim at optimizing for arithmetic intensity and SIMD vectorization, which obviously have a strong impact when arrays and iteration spaces are large. The corner cases of this phenomenon are indeed $[q=1, p=1]$ and $[q=4, p=4]$. We also observe how \texttt{auto}, in almost all scenarios, outperforms all of the other variants. In particular, it is not a surprise that \texttt{auto} is faster than \texttt{fix}, since \texttt{fix} is one of the autotuner's tested variants, as explained in Section~\ref{sec:autotune}. This proves the quality of the work presented in this paper, which shows significant advances over~\cite{Luporini}. The reasons for which \texttt{auto} exceeds both original code and \texttt{ffc} are discussed for each specific problem next. Also, details on the ``optimal'' code variant determined by autotuning are given in Section~\ref{sec:auto-analysis}.

\paragraph{Helmholtz}
\begin{figure}[t]  
\includegraphics[scale=0.7]{perf-results/helmholtz}
\caption{Helmholtz results.}\label{fig:helmholtz}
\end{figure}
The results for the Helmholtz problem are provided in Figure~\ref{fig:helmholtz}. We observe that \texttt{ffc} slows the code down, especially for $q \geq 3$. This is a consequence of using indirection arrays in the generated code that, as explained in Section~\ref{sec:zeros}, prevent, among the other compiler's optimizations, SIMD auto-vectorization. The \texttt{auto} version results in minimal performance improvements over \texttt{fix} when $nf=0$, unless $q=4$. This is due to the fact that if the loop over quadrature points is relatively small, then close-to-peak performance is obtainable through basic expression rewriting and code specialization; in this circumstance, generalized loop-invariant code motion and padding plus data alignment. The trend changes dramatically as $nf$ and $q$ increase: a more ample spectrum of transformations must be considered to find the optimal local assembly implementation. We will provide details about the selected transformations in the next section.

\paragraph{Elasticity}
\begin{figure}[t]  
\includegraphics[scale=0.7]{perf-results/elasticity}
\caption{Elasticity results.}\label{fig:elasticity}
\end{figure}
Figure~\ref{fig:elasticity} illustrates results for the Elasticity problem. This form uses a vector-valued space for the basis functions, so here transformations avoiding computation over zero-valued columns are of key importance. The \texttt{ffc} set of optimizations leads to notable improvements over the original code at $q=1$. The use of inderection arrays allows to phisically eliminate zero-valued columns at code generation time; as a consequence, different tabulated basis functions are merged into a single array. Therefore, despite the execution being purely scalar because of indirection arrays, the reduction in arithmetic intensity and register pressure imply improvement in performance. Nevertheless, \texttt{auto} remains in general the best choice, with gains over \texttt{ffc} that are wider as $p$ and $nf$ increase.

For $q \geq 2$, in \texttt{ffc} the lack of SIMD vectorization counterbalances the decrease in the number of floating point operations, leading to speed ups over the original code that only occasionally exceed 1$\times$. On the other hand, the successful application of the zero-avoidance optimization while preserving code specialization plays a key role for \texttt{auto}, resulting in much higher performance code especially at $q=2$ and $q=3$.

@@

Speed ups for the hyperelasticity form are shown in Figure~\ref{fig:hyperelasticity}. Experiments for $nf \geq 2$ could not be executed because of FEniCS-Form-Compiler's technical limitations.
